---
title: "Data Provenance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Provenance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(co2e)
```

## Overview

`co2e` bundles three lookup tables (`gwp_values`, `gases`, `gas_synonyms`),
built from a small set of source files in `data-raw/`. This vignette
documents where each value comes from and the choices made assembling them,
so a user checking a specific GWP figure against a published table can
trace it back to its source.

## Sources

| File | Contents | Source |
|---|---|---|
| `gwp_shortlist_epa.csv` | GWP values by gas, IPCC Assessment Report, and time horizon | EPA/data.gov IPCC AR4-AR6 GWPs (Federal LCA Commons), [doi:10.23719/1529821](https://doi.org/10.23719/1529821); AR6 methane fossil/non-fossil split rows from the GHG Protocol's *Global Warming Potential Values* (August 2024) |
| `gas_identity.csv` | Chemical formula and CAS Registry Number per gas | IPCC AR6 WGI Chapter 7 Supplementary Material, Table 7.SM.7 (via the chrisroadmap/ar6 CSV mirror); `gas_group` assigned manually |
| `gas_synonyms_seed.csv` | Documented alternate names per gas | GHG Protocol *Global Warming Potential Values* (August 2024) footnotes, IPCC AR6 Table 7.SM.7, and EPA/data.gov naming |

`gas_synonyms` also includes each gas's EPA/data.gov name, added
automatically from `gwp_shortlist_epa.csv`'s `epa_name` column during the
build (see `data-raw/build_data.R`).

## AR6 values are taken from the final published tables

IPCC AR6 WGI Chapter 7 Supplementary Material, Table 7.SM.7, as hosted on 
ipcc.ch, contains incorrect values for CFC-11 and CFC-12. This is acknowledged 
by the chapter's lead author for the metrics section (Bill Collins) via the 
chapter's supplementary GitHub repository, but as of this writing no formal 
IPCC corrigendum has been issued. co2e uses the corrected values (sourced from
the chapter author's repository and independently corroborated by the GHG 
Protocol's Global Warming Potential Values, August 2024) rather than the 
literal printed table for these two substances. This is the only known 
discrepancy between co2e and the printed Table 7.SM.7.

## Scope

- **Single substances only.** `gases$is_blend` is currently always `FALSE`
  - refrigerant blends (e.g. R-410A) and ASHRAE R-number naming generally
  are not yet in scope. `standardize_gas()` will return `"unmatched"` for
  R-number names like `"R-134a"` until blend support is added.
- **Three time horizons.** Only the 20-, 100-, and 500-year GWP horizons
  are included, matching what the source tables report.

## Reproducing the built data

The tables shipped in `data/` are built from the CSVs in `data-raw/` by
`data-raw/build_data.R`. Run it from the package root to regenerate them
after editing a source CSV:

```r
source("data-raw/build_data.R")
```
